---
title: "Practical 5"
author: "Ana Martins"
date: "2022-12-15"
output: html_document
---

## Introduction

```{r}
library(MASS)
library(tidyverse)
library(patchwork)
library(ggdendro)
```

**1. The code does not have comments. Add descriptive comments to the code below.**

```{r}
# setting the random seed so the same results can be reproduced
set.seed(123)

# creating a 2x2 matrix 
sigma      <- matrix(c(1, .5, .5, 1), 2, 2)
# creating a random sample of 2 x 100 values with mean on 5
sim_matrix <- mvrnorm(n = 100, mu = c(5, 5), 
                      Sigma = sigma)
# setting the column names of this matrix
colnames(sim_matrix) <- c("x1", "x2")

# creating a tibble from the sim_matrix and assigning 3 random classes to the data
sim_df <- 
  sim_matrix %>% 
  as_tibble() %>%
  mutate(class = sample(c("A", "B", "C"), size = 100, 
                        replace = TRUE))

# setting the x2 and x1 values to a variation depending on the class
sim_df_small <- 
  sim_df %>%
  mutate(x2 = case_when(class == "A" ~ x2 + .5,
                        class == "B" ~ x2 - .5,
                        class == "C" ~ x2 + .5),
         x1 = case_when(class == "A" ~ x1 - .5,
                        class == "B" ~ x1 - 0,
                        class == "C" ~ x1 + .5))
# same thing as before but with a bigger variation
sim_df_large <- 
  sim_df %>%
  mutate(x2 = case_when(class == "A" ~ x2 + 2.5,
                        class == "B" ~ x2 - 2.5,
                        class == "C" ~ x2 + 2.5),
         x1 = case_when(class == "A" ~ x1 - 2.5,
                        class == "B" ~ x1 - 0,
                        class == "C" ~ x1 + 2.5))
```

2. Prepare two unsupervised datasets by removing the class feature.

```{r}
sim_df_small <-
  sim_df_small %>% 
  select(-class)

sim_df_large <-
  sim_df_large %>% 
  select(-class)
```

**3. For each of these datasets, create a scatterplot. Combine the two plots into a single frame (look up the patchwork package to see how to do this!) What is the difference between the two datasets?**

```{r}
sim_df_small_plot <-
  sim_df_small %>% 
  ggplot(mapping = aes(x1, x2)) +
  geom_point(color = "#1b9e77") +
  theme_minimal() +
  ggtitle("sim_df_small")

sim_df_large_plot <-
  sim_df_large %>% 
  ggplot(mapping = aes(x1, x2)) +
  geom_point(color = "#d95f02") +
  theme_minimal() +
  ggtitle("sim_df_large")

sim_df_small_plot + sim_df_large_plot
```

For the sim_df_large there is much more variation within the classes (you can actually see classes even though they come from random data).

## 2 Hierarchical clustering

**4. Run a hierarchical clustering on these datasets and display the result as dendrograms. Use euclidian distances and the complete agglomeration method. Make sure the two plots have the same y-scale. What is the difference between the dendrograms? (Hint: functions you’ll need are hclust, ggdendrogram, and ylim)**

```{r}
sim_df_small_dist <- dist(sim_df_small, method = "euclidean")
sim_df_large_dist <- dist(sim_df_large, method = "euclidean")

sim_df_small_hclust <- hclust(sim_df_small_dist, method = "complete")
sim_df_large_hclust <- hclust(sim_df_large_dist, method = "complete")

ggdendrogram(sim_df_small_hclust) +
  ylim(0, 8) +
  ggtitle("sim_df_small")
ggdendrogram(sim_df_large_hclust) +
  ylim(0, 8) +
  ggtitle("sim_df_large")
```

The one for the less variations has much less distance between "types" of clusters, i.e. the vertical distances are much smaller.

**5. For the dataset with small differences, also run a complete agglomeration hierarchical cluster with manhattan distance.**

```{r}
sim_df_small_dist_manhattan <- dist(sim_df_small, method = "manhattan")
sim_df_small_manhattan_hclust <- hclust(sim_df_small_dist_manhattan, method = "complete")
```

**6. Use the cutree() function to obtain the cluster assignments for three clusters and compare the cluster assignments to the 3-cluster euclidian solution. Do this comparison by creating two scatter plots with cluster assignment mapped to the colour aesthetic. Which difference do you see?**

```{r}
sim_df_small_cut <- cutree(sim_df_small_hclust, k = 3)
sim_df_small_manhattan_cut <- cutree(sim_df_small_manhattan_hclust, k = 3)

sim_df_small <-
  sim_df_small %>% 
  mutate(class = sim_df_small_cut, class_manhattan = sim_df_small_manhattan_cut)

sim_df_small_clustered_plot <-
  sim_df_small %>%
  ggplot(mapping = aes(x1, x2, color = as.factor(class))) +
  geom_point() +
  theme_minimal() +
  ggtitle("euclidean")

sim_df_small_manhattan_clustered_plot <-
  sim_df_small %>% 
  ggplot(mapping = aes(x1, x2, color = as.factor(class_manhattan))) +
  geom_point() + 
  theme_minimal() +
  ggtitle("manhattan")

sim_df_small_clustered_plot + sim_df_small_manhattan_clustered_plot
```

The one with the manhattan distance divides the points into 2 different clusters even with points very close to each other.

## 3 K-means clustering

**7. Create k-means clusterings with 2, 3, 4, and 6 classes on the large difference data. Again, create coloured scatter plots for these clusterings.**

```{r}
sim_df_large_2 <- kmeans(sim_df_large, 2)$cluster
sim_df_large_3 <- kmeans(sim_df_large, 3)$cluster
sim_df_large_4 <- kmeans(sim_df_large, 4)$cluster
sim_df_large_6 <- kmeans(sim_df_large, 6)$cluster

sim_df_large <-
  sim_df_large %>% 
  mutate(cluster2 = sim_df_large_2, cluster3 = sim_df_large_3, cluster4 = sim_df_large_4, cluster6 = sim_df_large_6)

sim_df_large %>% 
  ggplot(mapping = aes(x1, x2, color = as.factor(cluster2))) +
  geom_point() +
  theme_minimal() + 
  ggtitle("2 clusters")

sim_df_large %>% 
  ggplot(mapping = aes(x1, x2, color = as.factor(cluster3))) +
  geom_point() +
  theme_minimal() + 
  ggtitle("3 clusters")

sim_df_large %>% 
  ggplot(mapping = aes(x1, x2, color = as.factor(cluster4))) +
  geom_point() +
  theme_minimal() + 
  ggtitle("4 clusters")

sim_df_large %>% 
  ggplot(mapping = aes(x1, x2, color = as.factor(cluster6))) +
  geom_point() +
  theme_minimal() + 
  ggtitle("6 clusters")
```

**8. Do the same thing again a few times. Do you see the same results every time? where do you see differences?**

```{r}
sim_df_large_2 <- kmeans(sim_df_large, 2)$cluster
sim_df_large_3 <- kmeans(sim_df_large, 3)$cluster
sim_df_large_4 <- kmeans(sim_df_large, 4)$cluster
sim_df_large_6 <- kmeans(sim_df_large, 6)$cluster

sim_df_large <-
  sim_df_large %>% 
  mutate(cluster2 = sim_df_large_2, cluster3 = sim_df_large_3, cluster4 = sim_df_large_4, cluster6 = sim_df_large_6)

sim_df_large %>% 
  ggplot(mapping = aes(x1, x2, color = as.factor(cluster2))) +
  geom_point() +
  theme_minimal() + 
  ggtitle("2 clusters")

sim_df_large %>% 
  ggplot(mapping = aes(x1, x2, color = as.factor(cluster3))) +
  geom_point() +
  theme_minimal() + 
  ggtitle("3 clusters")

sim_df_large %>% 
  ggplot(mapping = aes(x1, x2, color = as.factor(cluster4))) +
  geom_point() +
  theme_minimal() + 
  ggtitle("4 clusters")

sim_df_large %>% 
  ggplot(mapping = aes(x1, x2, color = as.factor(cluster6))) +
  geom_point() +
  theme_minimal() + 
  ggtitle("6 clusters")
```

The differences are mostly on the methods with 2 and 4 clusters.

**9. Find a way online to perform bootstrap stability assessment for the 3 and 6-cluster solutions.**

```{r}
library(fpc)

sim_df_large_assessment3 <- clusterboot(sim_df_large,
                                        clustermethod = hclustCBI,
                                        method = "ward",
                                        k = 3)

sim_df_large_assessment3$bootmean
sim_df_large_assessment3$bootbrd

sim_df_large_assessment6 <- clusterboot(sim_df_large,
                                        clustermethod = hclustCBI,
                                        method = "ward",
                                        k = 6)

sim_df_large_assessment6$bootmean
sim_df_large_assessment6$bootbrd
```

Clusters with higher bootbrd are more unstable.

## 4 Challenge question

**10. Create a function to perform k-medians clustering**

```{r}
kmedians <- function(data, k) {
  loop <- TRUE
  prev <- c(rep(0, nrow(data)))
  
  it <- 0
  
  while (loop) {
    s <- c(1:k)
    
    cluster <-
      sample(s, nrow(data), replace = TRUE, prob = c(rep(100 / k, k)))
    
    data <-
      data %>%
      select(x1, x2) %>%
      mutate(cluster = cluster)
    
    for (i in 1:k) {
      datai <-
        data %>%
        filter(cluster == i)
      
      x[i] <- median(datai$x1)
      y[i] <- median(datai$x2)
      
    }
    
    cl <- c(rep(1, nrow(data)))
    
    for (i in 1:nrow(data)) {
      smallest <- sqrt((x[1] - data$x1[i]) ^ 2 + (y[1] - data$x2[i]) ^ 2)
      for (j in 2:k) {
        if (sqrt((x[j] - data$x1[i]) ^ 2 + (y[j] - data$x2[i]) ^ 2) < smallest) {
          smallest <- sqrt((x[j] - data$x1[i]) ^ 2 + (y[j] - data$x2[i]) ^ 2)
          cl[i] <- j
        }
      }
      
    }
    
    data <-
      data %>%
      mutate(cluster = cl)
    
    sum <- 0
    
    for (i in 1:nrow(data)) {
      if (prev[i] == data$cluster[i]) {
        sum <- sum + 1
      }
    }
    
    if (sum == 100)
      loop <- FALSE
    
    prev <- data$cluster
    it <- it + 1
    
  }
  
  print(it)
  
  prev
  
}

out <- kmedians(sim_df_large, 3)
```

**11. Add an input parameter smart_init. If this is set to TRUE, initialize cluster assignments using hierarchical clustering (from hclust). Using the unsupervised sim_df_small, look at the number of iterations needed when you use this method vs when you randomly initialize.**

```{r}
kmedians <- function(data, k, start_init = FALSE) {
  loop <- TRUE
  prev <- c(rep(0, nrow(data)))
  
  it <- 0
  
  while (loop) {
    s <- c(1:k)
    
    if (start_init) {
      dist <- dist(data)
      hclust <- hclust(dist)
      cluster <- cutree(hclust, k)
    } else {
      cluster <-
        sample(s,
               nrow(data),
               replace = TRUE,
               prob = c(rep(100 / k, k)))
    }
    
    data <-
      data %>%
      select(x1, x2) %>%
      mutate(cluster = cluster)
    
    for (i in 1:k) {
      datai <-
        data %>%
        filter(cluster == i)
      
      x[i] <- median(datai$x1)
      y[i] <- median(datai$x2)
      
    }
    
    cl <- c(rep(1, nrow(data)))
    
    for (i in 1:nrow(data)) {
      smallest <- sqrt((x[1] - data$x1[i]) ^ 2 + (y[1] - data$x2[i]) ^ 2)
      for (j in 2:k) {
        if (sqrt((x[j] - data$x1[i]) ^ 2 + (y[j] - data$x2[i]) ^ 2) < smallest) {
          smallest <- sqrt((x[j] - data$x1[i]) ^ 2 + (y[j] - data$x2[i]) ^ 2)
          cl[i] <- j
        }
      }
      
    }
    
    data <-
      data %>%
      mutate(cluster = cl)
    
    sum <- 0
    
    for (i in 1:nrow(data)) {
      if (prev[i] == data$cluster[i]) {
        sum <- sum + 1
      }
    }
    
    if (sum == 100)
      loop <- FALSE
    
    prev <- data$cluster
    it <- it + 1
    
  }
  
  print(it)
  
  prev
  
}

out <- kmedians(sim_df_large, 3, start_init = TRUE)
```

