---
title: 'Practical 8: Text Mining'
author: "Ana Martins"
date: "2023-01-13"
output: html_document
---

## 1 Introduction

```{r}
library(tidyverse)
library(tidytext)
library(tm)
library(e1071)
library(topicmodels)
library(stringi)
```

## 2 Vector space model: document-term matrix

**1. Use the code below to load the data set and inspect its first rows.**

```{r}
load("data/news_dataset.rda")
head(df_final)
```

**2. Find out about the name of the categories and the number of observations in each of them.**

```{r}
df_final <-
  df_final %>% 
  mutate(Category = as.factor(Category))

summary(df_final$Category)
```

**3. Convert the data set into a document-term matrix and use the findFreqTerms function to keep the terms which their frequency is higher than 10. It is also a good idea to apply some text preprocessing before this conversion: e.g., remove non-UTF-8 characters, convert the words into lowercase, remove punctuation, numbers, stopwords, and whitespaces.**

```{r}
dtm <- DocumentTermMatrix(df_final, control = list(removePunctuation = TRUE, removeNumbers = TRUE, stopwords = TRUE))
inspect(dtm)
```

